Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-12-08 paper.tex
Add structure for conclusion section
Matt's edits through p. 14: warn about sloppy treatment of flavor ev

In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present an application
in the dynamical environment following the merger of two neutron stars,
computing neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.

@@

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
In this way we approximate the system as stationary over a light-crossing time
of around 1~ms, which is far below the thermal timescale of the remnant.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
\footnote{TODO: way too-low angular resolution}
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.

@@

\subsection{Neutrino Oscillations at High Neutrino Densities}
\label{ssec:V_nunu}
The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations for neutrinos
\cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths, i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions.
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.

But at the opposite limit, in the free-streaming regime,
coherence effects become important,
and the quantum kinetic equations take a Schr\"{o}dinger-like form:
\begin{equation}
  \label{eqn:schroedinger}
  i \frac{d}{ds} S = H S,
\end{equation}
where $ds$ is an interval of proper length, and
\footnote{TODO: mention $\bar{\nu}$ eqn}
$S$ is the flavor evolution matrix describing a mixed state
\begin{equation}
  \label{eq:density_evolution}
  \rho = S \, \rho_0 \, S^\dagger.
\end{equation}
In this limit, we may decompose the Hamiltonian into vacuum,
matter, and neutrino contributions:
\begin{equation}
  \label{eqn:decompose_H}
  H = H_{\rm V} + H_e + H_{\nu\nu}.
\end{equation}
Explicit formulas for these terms may be found in the oscillation literature,
e.g.\ \cite{duan2009-review},
but for clarity we give their order-of-magnitude scales here:
\begin{align}
  \label{eqn:scale_H_V}
  |H_{\rm V}| &\sim \frac{\Delta m^2}{\varepsilon}, \\
  \label{eqn:scale_H_e}
  |H_e|       &\sim G_{\rm F}\left|n_{e^-}-n_{e^+}\right|, \\
  \label{eqn:scale_H_nunu}
  |H_{\nu\nu}|&\sim G_{\rm F} \left|G_{\nu_e}-G_{\bar{\nu}_e}\right|,
\end{align}
where $\Delta m$ is the mass difference between neutrino mass eigenstates,
$G_{\rm F}$ is the Fermi coupling constant,
$n_{e^-}$ and $n_{e^+}$ are the electron and positron number densities,
and $G_{\nu_e}$ and $G_{\bar{\nu}_e}$ the electron neutrino and antineutrino
number densities defined in Eqn.~\ref{eqn:G}.
For the scale of $H_{\nu\nu}$ we have only included the isotropic
components of the neutrino fields.

@@

>>>>

In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present an application
in the dynamical environment following the merger of two neutron stars,
computing neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.
In Sec.~\ref{sec:conclusions} we summarize our work and anticipate improvements.

@@

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
In this way we approximate the system as stationary over a light-crossing time
of around 1~ms, which is far below the thermal timescale of the remnant.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.

@@

\subsection{Neutrino Oscillations at High Neutrino Densities}
\label{ssec:V_nunu}
In this section we examine the importance of elastic scattering in neutrino
flavor oscillation above the hypermassive neutron star--disk configuration
presented in Sec.~\ref{ssec:test_disk_comparison}.
Our treatment of the flavor evolution equation assumes flat space and small fluid
velocities.
In consequence we treat some of gauge-dependent quantities inconsistently,
and we ignore potentially important features of relativistic flavor evolution
near compact objects.
\footnote{TODO: cite Yang et Kneller}
Such a cursory treatment is sufficient, however, for the following qualitative
exploration.

The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations for neutrinos
\cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths, i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions.
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.

But at the opposite limit, in the free-streaming regime,
coherence effects become important,
and the quantum kinetic equations take a Schr\"{o}dinger-like form:
\begin{equation}
  \label{eqn:schroedinger}
  i \frac{d}{ds} S = H S,
\end{equation}
where $ds$ is an interval of proper length traversed by the neutrino as
measured by our observer, and
\footnote{TODO: mention $\bar{\nu}$ eqn}
$S$ is the flavor evolution matrix describing a mixed state
\begin{equation}
  \label{eq:density_evolution}
  \rho = S \, \rho_0 \, S^\dagger.
\end{equation}
In this limit, we may decompose the Hamiltonian into vacuum,
matter, and neutrino contributions:
\begin{equation}
  \label{eqn:decompose_H}
  H = H_{\rm V} + H_e + H_{\nu\nu}.
\end{equation}
Explicit formulas for these terms may be found in the oscillation literature,
e.g.\ \cite{duan2009-review},
but for clarity we give their order-of-magnitude scales here:
\begin{align}
  \label{eqn:scale_H_V}
  |H_{\rm V}| &\sim \frac{\Delta m^2}{\varepsilon}, \\
  \label{eqn:scale_H_e}
  |H_e|       &\sim G_{\rm F}\left|n_{e^-}-n_{e^+}\right|, \\
  \label{eqn:scale_H_nunu}
  |H_{\nu\nu}|&\sim G_{\rm F} \left|G_{\nu_e}-G_{\bar{\nu}_e}\right|,
\end{align}
where $\Delta m$ is the mass difference between neutrino mass eigenstates,
$G_{\rm F}$ is the Fermi coupling constant,
$n_{e^-}$ and $n_{e^+}$ are the electron and positron number densities,
and $G_{\nu_e}$ and $G_{\bar{\nu}_e}$ the electron neutrino and antineutrino
number densities defined in Eqn.~\ref{eqn:G}.
For the scale of $H_{\nu\nu}$ we have only included the isotropic
components of the neutrino fields.

@@

\section{Conclusions}
\label{sec:conclusions}
Add some conclusions.
\footnote{TODO: note some improvements needed: implement implicit time-stepping,
  use finite-difference grid for fluid vars, improve gray spectral assumptions}
